---
title: "Zbaracki et al. (2021)"
---

Zbaracki, Mark J., Watkiss, Lee, McAlpine, Cameron and Barg, Julian (2021), "Truth, Beauty, and Justice in Models of Social Action", Beckman, Christine M. (Ed.) *Carnegie goes to California: Advancing and Celebrating the Work of James G. March (Research in the Sociology of Organizations, Vol. 76)*, Emerald Publishing Limited, Bingley, pp. 159-177. https://www.doi.org/10.1108/S0733-558X20210000076007

<font size="2">This author accepted manuscript is deposited under a Creative Commons Attribution Non-commercial 4.0 International (CC BY-NC) licence. This means that anyone may distribute, adapt, and build upon the work for non-commercial purposes, subject to full attribution. If you wish to use this manuscript for commercial purposes, please contact permissions@emerald.com.</font>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r CV, layout="l-screen-inset shaded"}
knitr::include_url("https://julianbarg.github.io/public_documents/Zbaracki%20et%20al%20(2021).pdf", heigh = "4500px")
```
