---
title: "How (the heck) does ggalluvial work?"
draft: false
author: Julian Barg
output:
  distill::distill_article:
    self_contained: false
---

We have this unfortunate problem that we want to create a sankey diagram, but
most of the existing packages are not feature complete enough. ggsankey is a 
popular solution, however, the sankey diagrams it produces leave much of the 
space unused when working with a large dataset and there are no customization
options to change this. Then, there is the networkD3, which is actually just an
R wrapper for the D3.js JavaScript library. Again, it inserts a lot of white 
whitespace and there are only some basic customization options. 

Fortunately, I came across ggalluvial, which makes much better use of space. 
Technically, that is because it makes different assumptions about the data. 
An alluvial plot tracks changes of attributes across subjects over time, whereas
a sankey diagram is designed to track track quantities. Because of this 
underlying assumption that there is a change of state between a and b, not a 
flow from a to b, we retain more control. For example, we can to some degree
control attributes such as color even after we have already gone from a to b,
for subsections of the cohort in b. The underlying data structure of a sankey
diagram does not typically track observations across different stages, so there
is no means of controlling attributes such as color once the train has left the
station.

However, the inner workings are still a mystery to me, so let's run some tests
to see how this whole thing works.

```{r library, echo = F}
library(tidyverse)
library(ggalluvial)
```

## Basic example

As you can see, the number of variables we need is quite high. x and y are 
quite straightforward. 

* `x` is simply a ordinal variable which denotes the
axes. 
* `stratum` then denotes that value of the observations on that 
particular axis. 
* `alluvium` allows for tracking observations across the 
axes.
* `y` allows us to set a count.

```{r }
alluvial <- tribble(
  ~x, ~stratum, ~alluvium, ~y, ~label, ~fill, ~flow,
  "first", "high", 1, 35, "high", "red", "yellow",
  "second", "high", 1, 35, "high", "red", "yellow",
  "first", "low", 2, 5, "low", "blue", "pink",
  "second", "low", 2, 10, "low", "blue", "pink"
)

alluvial %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = fill, y = y)) +
    geom_flow() +
    geom_stratum() +
    geom_text(aes(label = label), stat = "stratum")
```

The geoms do not respect the colors we selected for fill, so we have to go in
and fix that manually with `scale_fill_manual`. Still, this allow us to set
the color for flows and strata manually and independently.

### Reordering flows

One awkward attribute of this data structure is that the factor order of the 
factor variable `stratum` controls the order of variables across axes, so 
management of the factor levels is important and awkward.

```{r order}
alluvial %>%
  mutate(stratum = factor(stratum, c("low", "high"))) %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = fill, y = y)) +
    geom_flow(aes(fill = flow), color = "black") +
    geom_stratum() +
    geom_text(aes(label = label), stat = "stratum")
```

### Joining observations

We can merge different observations and then have them split off again. 
Importantly, this impairs our ability to set colors. Notice how `medium` is 
colored in in red, even though there should be blue in there, too. So the much
better option is to use `stratum` directly to control fill.

```{r }
alluvial_merge <- tribble(
  ~x, ~stratum, ~alluvium, ~y, ~label, ~fill, ~flow,
  "first", "high", 1, 35, "high", "red", "yellow",
  "second", "medium", 1, 35, "medium", "red", "yellow",
  "third", "high", 1, 35, "high", "red", "yellow",
  "first", "low", 2, 10, "low", "blue", "pink",
  "second", "medium", 2, 10, "medium", "red", "pink",
  "third", "low", 2, 10, "low", "blue", "pink"
)
alluvial_merge %>%
  mutate(stratum = factor(stratum, c("low", "high"))) %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = fill, y = y)) +
    geom_flow(aes(fill = flow), color = "black") +
    geom_stratum() +
    geom_text(aes(label = label), stat = "stratum") +
    scale_fill_manual(values = c("blue" = "blue", "red" = "red", 
                                 "pink" = "pink", "yellow" = "yellow"))
```

Joined observations are treated as the same object if they share start and end
point after joining. 

```{r distinct}
alluvial_distinct <- tribble(
  ~x, ~stratum, ~alluvium, ~attribute,
  "first", "high", 1, "long", 
  "second", "medium", 1, "long",
  "third", "medium", 1, "long",
  "first", "low", 2, "short", 
  "second", "medium", 2, "short",
  "third", "medium", 2, "short"
)

alluvial_distinct %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = stratum)) +
    geom_flow(color = "black") +
    geom_stratum() +
    geom_text(aes(label = stratum), stat = "stratum")
```

That is a problem when we want to show in `geom_flow` any variable other than
the variables to the left and the right. We can use the `fill` argument to do 
so, but we will need to use the `after_stat` function to make sure it's 
colored in correctly.

```{r sep}
alluvial_distinct %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = attribute)) +
    geom_flow(color = "black") +
    geom_stratum(aes(fill = after_stat(stratum))) +
    geom_text(aes(label = stratum), stat = "stratum")
```

### Reordering

We need to figure out what determines the order of the inflow and outflow into 
the same stratum. Can we manipulate this somehow without changing the stratum?

```{r stratum}
stratum_control <- tribble(
  ~x, ~stratum, ~alluvium, ~attribute, ~label, ~trick,
  "first", "high", 1, "high", "high", "high",
  "second", "wide", 1, "high", "wide", "wide-low", 
  "third", "wide", 1, "high", "wide", "wide-low", 
  "first", "low", 2, "low", "low", "wide",
  "second", "wide", 2, "low", "wide", "wide",
  "third", "wide", 2, "low", "wide", "wide", 
)

stratum_control %>%
  mutate(stratum = factor(stratum, c("low", "medium", "high"))) %>%
  mutate(attribute = factor(attribute, c("low", "medium", "high"))) %>%
  mutate(trick = factor(trick, c("wide-low", "wide"))) %>%
  ggplot(aes(x = x, stratum = stratum, alluvium = alluvium, fill = attribute, 
             label = label)) +
    geom_flow(aes(stratum = trick), color = "black") +
    geom_stratum(aes(fill = after_stat(label))) +
    geom_text(stat = "stratum")
```

Fabulous, with this trick we can control flows independently from 
the axes.
